---
title: "RNAseq Differential Expression"
output:
  pdf_document: default
  html_notebook: default
---
--------



-----------
#installing bioconductor packages
```{r include=FALSE}
#install.packages("devtools")
#install.packages("BiocManager")
#BiocManager::install("TxDb.Rnorvegicus.UCSC.rn6.refGene")
BiocManager::install("Rsubread")
BiocManager::install("DESeq2")
BiocManager::install("pheatmap")
#BiocManager::install("Rattus.norvegicus")
BiocManager::install("EnhancedVolcano")
BiocManager::install("topGO")
BiocManager::install("gridExtra")
#BiocManager::install("org.Rn.eg.db")
BiocManager::install("DOSE")
BiocManager::install("clusterProfiler")
BiocManager::install("pathview")
#BiocManager::install("fgsea")
BiocManager::install("factoextra")
BiocManager::install("scRNAseq")
BiocManager::install("VennDiagram")
BiocManager::install("babelgene")
```

#Load libraries
```{r include=FALSE}
library(ggplot2)
library(rgl)
library(tidyr)
library(tidyverse)
library(RColorBrewer)
library(expss)
library(dplyr)
library(BiocManager)
library(Rsamtools)
library(GenomicAlignments)
library(BiocParallel)
library(GenomicFeatures)
library(Rsubread)
library(DESeq2)
library(Rsamtools)
library(pheatmap)
library(AnnotationDbi)
library(Mus.musculus)
library(EnhancedVolcano)
library(TxDb.Mmusculus.UCSC.mm10.knownGene)
library(topGO)
library(org.Mm.eg.db)
library(gridExtra)
library(DOSE)
library(clusterProfiler)
library(fgsea)
library(factoextra)
library(pathview)
library(enrichplot)
library(readr)
library(ggpubr)
library(ggsignif)
library(rstatix)
library(scRNAseq)
library(VennDiagram)
library(stringr)
library(tibble)
library(org.Hs.eg.db)
library(viridis)
library(readr)
library(patchwork)
library(babelgene)
```


#***Using DeSeq2***
We can now construct a DESeqDataSet object to formulate a starting point for our analysis. You need to add an appropriate design for analysis

#removing outliers identified to clean and rerun analysis
```{r}
se1<- se[,se$tissue_type == "Dorsal Hippocampus"] ##change tissue type here

se1<-se1[ , !se1$bam_ID == c("Li21.bam")] #remove specific outliers here

se1 <- se1[rowSums(assay(se1)) >= 10, ]

#M_se<-se1[,se1$sex == "male"]
#rm(F_se) #remove variable
#rm(se1) #remove variable
#test<-M_se #[ , !se1$animal_ID ==] #remove specific outliers here
#colData(test)

se1$group<-paste(se1$exposure, se1$sex, sep = "_") #create a group variable for easy comparison
se1$group

#se1 <- se1[ , order( colnames( se1 ) ) ]
#head(assay(se1))
#colData(se1)

se1$group<-factor(se1$group)
se1$group
dds <- DESeqDataSet(se1, 
                    design = ~ group) #comparison done between high treatment and vehicle?
```

```{r}
#dds$exposure<- factor(dds$exposure, levels = c("control","PCB"))

```


#***Exploratory analysis and visualization***
*transformations*
The variance stabilizing transformation (VST) a goal of stablizing the variance across the range of values. produce log2-like values for high counts. 
```{r}
colData(dds)

vsd <- vst(dds)
```


Now for visuals, plot a principal components analysis (PCA) using  ggplot2
```{r}
data <- plotPCA(vsd, 
                  intgroup = 'group' , 
                  returnData=TRUE)
data
percentVar <- round(100 * attr(data, 
                                 "percentVar"))
```
#PCA Plot
```{r}
tiff(file = 'Figure/PCA/Hippocampus_PCA.tiff', units="in", width=8, height=5, res=1000)
PCA <-ggplot(data, aes(PC1, 
                   PC2, 
                   color=group,
                   shape= group)) + 
  stat_ellipse(aes(color= group))+
  ggtitle("Hippocampus PCA")+
  geom_point(size=3) +
  scale_color_manual(values = c("#666666", "#FF66CC","#66CCCC", "#9966FF")) +
  scale_shape_manual(values = c(19, 15,17, 18)) +
  xlab(paste0("PC1: ",
              percentVar[1],
              "% variance")) +
  ylab(paste0("PC2: ",
              percentVar[2],
              "% variance"))+ 
  theme_bw() + 
  theme(panel.border =element_rect(colour = "black", fill=NA, size=1), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"))
PCA
dev.off()
PCA
```

#***Differential Expression***
Lets run a differential expression pipeline using the raw counts

This function will print out a message for the various steps it performs. These are described in more detail in the manual page for DESeq, which can be accessed by typing ?DESeq. Briefly these are: the estimation of size factors (controlling for differences in the sequencing depth of the samples), the estimation of dispersion values for each gene, and fitting a generalized linear model.
```{r}
dds <- DESeq(dds)
resultsNames(dds)
```
Calling results without any arguments will extract the estimated log2 fold changes and p values for the last variable in the design formula. If there are more than 2 levels for this variable, results will extract the results table for a comparison of the last level over the first level. Treatment 6 vs vehicle
```{r}
res_exp<-results(dds, name = "exposure_PCB_vs_control")

res_exp

summary(res_exp)
```


#***Annotating results***
```{r}
columns(Mus.musculus)
```

How to assign actual gene names to our counts. Using an annotation package for mouse 
```{r}
res_exp$symbol <- mapIds(Mus.musculus, #which annotation package model to use
                     keys=row.names(res_exp), #using rownames
                     column="SYMBOL", #add a column of symbols associated with row.names arguments for the model
                     keytype="ENSEMBL", #our rownames are gene id's 
                     multiVals="first")
```

```{r}
res_exp$ONTOLOGY <- mapIds(Mus.musculus,
                     keys=row.names(res_exp),
                     column="ONTOLOGYALL",
                     keytype="ENSEMBL",
                     multiVals="first")
```

```{r}
res_exp$Genename <- mapIds(Mus.musculus,
                     keys=row.names(res_exp),
                     column="GENENAME", #now add a column for gene names or gene description
                     keytype="ENSEMBL",
                     multiVals="first")
```


```{r}
res_exp$ensembl <- mapIds(Mus.musculus,
                     keys=row.names(res_exp),
                     column="ENSEMBL", #now add a column for gene names or gene description
                     keytype="ENSEMBL",
                     multiVals="first")
```
```{r}
res_exp$ENTREZID <- mapIds(Mus.musculus,
                     keys=row.names(res_exp),
                     column="ENTREZID", #now add a column for gene names or gene description
                     keytype="ENSEMBL",
                     multiVals="first")
```


#***Plotting results***
Heatmaps are a good way to visualize the most significant genes
```{r}
mat <- assay(vsd)[ head(order(res_exp$padj),
                            30), #top 30 genes 
                       ]
#mat <- mat - rowMeans(mat)
df <- as.data.frame(colData(vsd)[,
                                     "exposure"])
rownames(df)<-colnames(mat)
colnames(df)[1]<- "Exposure"
```

```{r}
tiff(file = "Figure/Male_Liver_heatmap.tiff", units="in", width=8, height=5, res=1000)
heatmap<-pheatmap(mat, 
         annotation_col=df,
         labels_row = res_exp$symbol,
         fontsize = 6,
         scale = "row",
         show_colnames = F,
         cluster_cols = TRUE,
         main = "Male Liver Heatmap")
heatmap
dev.off()
heatmap
```

#***volcano plots***
```{r}
tiff(file = 'Figure/volcano/Male_Liver_volcano_plot.tiff', units="in", width=8, height=5, res=1000)
M_vol1<- EnhancedVolcano(res_exp,
    lab = res_exp$symbol,
    x = 'log2FoldChange',
    y = 'pvalue',
    title = "Male Liver Volcano Plot",
    subtitle = "",
    ylim = c(0, -log10(10e-8)),
    xlim = c(log2(10e-3), -log2(10e-3)),
    FCcutoff = 1.0,
    pCutoff = 0.05,
    labSize = 3.0,
    colAlpha = 1,
    legendLabels=c('Not sig.','Log (base 2) FC','p-value',
      'p-value & Log (base 2) FC'),
    legendPosition = 'right',
    legendLabSize = 12,
    legendIconSize = 3.0,
    gridlines.major = FALSE,
    gridlines.minor = FALSE,
    border = 'full',
    borderWidth = 0.5,
    borderColour = 'black')
M_vol1
dev.off()

M_vol1
```
#Export TSV results

```{r}
exp_res <- as.data.frame(res_exp)
write_tsv(exp_res, file="male_liver_results.tsv")
write_csv(exp_res, file="male_liver_results.csv")
write_csv(as.data.frame(assay(M_se)), file= "male_liver_counts.csv")
```


#REPEAT FOR FEMALES
```{r}
se1<- se[,se$tissue_type == "Liver"] ##change tissue type here

se1<-se1[ , !se1$bam_ID == c("Li21.bam")] #remove specific outliers here

se1 <- se1[rowSums(assay(se1)) >= 10, ]

F_se<-se1[,se1$sex == "female"]

#test<-M_se #[ , !se1$animal_ID ==] #remove specific outliers here
#colData(test)

se1$group<-paste(se1$exposure, se1$sex, sep = "_") #create a group variable for easy comparison
se1$group

#se1 <- se1[ , order( colnames( se1 ) ) ]
#head(assay(se1))
#colData(se1)

se1$group<-factor(se1$group)
se1$group
dds <- DESeqDataSet(F_se, 
                    design = ~ exposure) #comparison done between high treatment and vehicle?
```


#***Exploratory analysis and visualization***
*transformations*
The variance stabilizing transformation (VST) a goal of stablizing the variance across the range of values. produce log2-like values for high counts. 
```{r}
colData(dds)

vsd <- vst(dds)
```


Now for visuals, plot a principal components analysis (PCA) using  ggplot2
```{r}
data <- plotPCA(vsd, 
                  intgroup = 'exposure' , 
                  returnData=TRUE)
data
percentVar <- round(100 * attr(data, 
                                 "percentVar"))
```
#PCA Plot
```{r}
tiff(file = 'Figure/PCA/Female_Liver_PCA_outlier_removed.tiff', units="in", width=8, height=5, res=1000)
PCA <-ggplot(data, aes(PC1, 
                   PC2, 
                   color=group,
                   shape= group)) + 
  stat_ellipse(aes(color= group))+
  ggtitle("Female Liver PCA")+
  geom_point(size=3) +
  scale_color_manual(values = c("#666666", "#FF66CC","#66CCCC", "#9966FF")) +
  scale_shape_manual(values = c(19, 15,17, 18)) +
  xlab(paste0("PC1: ",
              percentVar[1],
              "% variance")) +
  ylab(paste0("PC2: ",
              percentVar[2],
              "% variance"))+ 
  theme_bw() + 
  theme(panel.border =element_rect(colour = "black", fill=NA, size=1), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"))
PCA
dev.off()
PCA
```

#***Differential Expression***
Lets run a differential expression pipeline using the raw counts

This function will print out a message for the various steps it performs. These are described in more detail in the manual page for DESeq, which can be accessed by typing ?DESeq. Briefly these are: the estimation of size factors (controlling for differences in the sequencing depth of the samples), the estimation of dispersion values for each gene, and fitting a generalized linear model.
```{r}
dds <- DESeq(dds)
resultsNames(dds)
```
Calling results without any arguments will extract the estimated log2 fold changes and p values for the last variable in the design formula. If there are more than 2 levels for this variable, results will extract the results table for a comparison of the last level over the first level. Treatment 6 vs vehicle
```{r}
res_exp<-results(dds, name = "exposure_PCB_vs_control")

res_exp

summary(res_exp)
```


#***Annotating results***
```{r}
columns(Mus.musculus)
```

How to assign actual gene names to our counts. Using an annotation package for mouse 
```{r}
res_exp$symbol <- mapIds(Mus.musculus, #which annotation package model to use
                     keys=row.names(res_exp), #using rownames
                     column="SYMBOL", #add a column of symbols associated with row.names arguments for the model
                     keytype="ENSEMBL", #our rownames are gene id's 
                     multiVals="first")
```

```{r}
res_exp$ONTOLOGY <- mapIds(Mus.musculus,
                     keys=row.names(res_exp),
                     column="ONTOLOGYALL",
                     keytype="ENSEMBL",
                     multiVals="first")
```

```{r}
res_exp$Genename <- mapIds(Mus.musculus,
                     keys=row.names(res_exp),
                     column="GENENAME", #now add a column for gene names or gene description
                     keytype="ENSEMBL",
                     multiVals="first")
```


```{r}
res_exp$ensembl <- mapIds(Mus.musculus,
                     keys=row.names(res_exp),
                     column="ENSEMBL", #now add a column for gene names or gene description
                     keytype="ENSEMBL",
                     multiVals="first")
```
```{r}
res_exp$ENTREZID <- mapIds(Mus.musculus,
                     keys=row.names(res_exp),
                     column="ENTREZID", #now add a column for gene names or gene description
                     keytype="ENSEMBL",
                     multiVals="first")
```


#***Plotting results***
Heatmaps are a good way to visualize the most significant genes
```{r}
mat <- assay(vsd)[ head(order(res_exp$padj),
                            30), #top 30 genes 
                       ]
#mat <- mat - rowMeans(mat)
df <- as.data.frame(colData(vsd)[,
                                     "exposure"])
rownames(df)<-colnames(mat)
colnames(df)[1]<- "Exposure"
```

```{r}
tiff(file = "Figure/Female_Liver_heatmap.tiff", units="in", width=8, height=5, res=1000)
heatmap<-pheatmap(mat, 
         annotation_col=df,
         labels_row = res_exp$symbol,
         fontsize = 6,
         scale = "row",
         show_colnames = F,
         cluster_cols = TRUE,
         main = "Female Liver Heatmap")
heatmap
dev.off()
heatmap
```

#***volcano plots***
```{r}
tiff(file = 'Figure/volcano/Female_Liver_volcano_plot.tiff', units="in", width=8, height=5, res=1000)
M_vol1<- EnhancedVolcano(res_exp,
    lab = res_exp$symbol,
    x = 'log2FoldChange',
    y = 'pvalue',
    title = "Female Liver Volcano Plot",
    subtitle = "",
    ylim = c(0, -log10(10e-8)),
    xlim = c(log2(10e-3), -log2(10e-3)),
    FCcutoff = 1.0,
    pCutoff = 0.05,
    labSize = 3.0,
    colAlpha = 1,
    legendLabels=c('Not sig.','Log (base 2) FC','p-value',
      'p-value & Log (base 2) FC'),
    legendPosition = 'right',
    legendLabSize = 12,
    legendIconSize = 3.0,
    gridlines.major = FALSE,
    gridlines.minor = FALSE,
    border = 'full',
    borderWidth = 0.5,
    borderColour = 'black')
M_vol1
dev.off()

M_vol1
```
#Export TSV results

```{r}
exp_res <- as.data.frame(res_exp)
write_tsv(exp_res, file="female_liver_results.tsv")
write_csv(exp_res, file="female_liver_results.csv")
write_csv(as.data.frame(assay(M_se)), file= "female_liver_counts.csv")
```

#Repeat for Hippocampus Male
```{r}
se1<- se[,se$tissue_type == "Dorsal Hippocampus"] ##change tissue type here

#se1<-se1[ , !se1$bam_ID == c("Li21.bam")] #remove specific outliers here

se1 <- se1[rowSums(assay(se1)) >= 10, ]

M_se<-se1[,se1$sex == "male"]
#rm(F_se) #remove variable
#rm(se1) #remove variable
#test<-M_se #[ , !se1$animal_ID ==] #remove specific outliers here
#colData(test)

se1$group<-paste(se1$exposure, se1$sex, sep = "_") #create a group variable for easy comparison
se1$group

#se1 <- se1[ , order( colnames( se1 ) ) ]
#head(assay(se1))
#colData(se1)

se1$group<-factor(se1$group)
se1$group
dds <- DESeqDataSet(M_se, 
                    design = ~ exposure) #comparison done between high treatment and vehicle?
```


#***Exploratory analysis and visualization***
*transformations*
The variance stabilizing transformation (VST) a goal of stablizing the variance across the range of values. produce log2-like values for high counts. 
```{r}
colData(dds)

vsd <- vst(dds)
```


Now for visuals, plot a principal components analysis (PCA) using  ggplot2
```{r}
data <- plotPCA(vsd, 
                  intgroup = 'exposure' , 
                  returnData=TRUE)
data
percentVar <- round(100 * attr(data, 
                                 "percentVar"))
```
#PCA Plot
```{r}
tiff(file = 'Figure/PCA/Male_Dorsal_Hippocampus_PCA_outlier_removed.tiff', units="in", width=8, height=5, res=1000)
PCA <-ggplot(data, aes(PC1, 
                   PC2, 
                   color=group,
                   shape= group)) + 
  stat_ellipse(aes(color= group))+
  ggtitle("Male Dorsal Hippocampus PCA")+
  geom_point(size=3) +
  scale_color_manual(values = c("#666666", "#FF66CC","#66CCCC", "#9966FF")) +
  scale_shape_manual(values = c(19, 15,17, 18)) +
  xlab(paste0("PC1: ",
              percentVar[1],
              "% variance")) +
  ylab(paste0("PC2: ",
              percentVar[2],
              "% variance"))+ 
  theme_bw() + 
  theme(panel.border =element_rect(colour = "black", fill=NA, size=1), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"))
PCA
dev.off()
PCA
```

#***Differential Expression***
Lets run a differential expression pipeline using the raw counts

This function will print out a message for the various steps it performs. These are described in more detail in the manual page for DESeq, which can be accessed by typing ?DESeq. Briefly these are: the estimation of size factors (controlling for differences in the sequencing depth of the samples), the estimation of dispersion values for each gene, and fitting a generalized linear model.
```{r}
dds <- DESeq(dds)
resultsNames(dds)
```
Calling results without any arguments will extract the estimated log2 fold changes and p values for the last variable in the design formula. If there are more than 2 levels for this variable, results will extract the results table for a comparison of the last level over the first level. Treatment 6 vs vehicle
```{r}
res_exp<-results(dds, name = "exposure_PCB_vs_control")

res_exp

summary(res_exp)
```


#***Annotating results***
```{r}
columns(Mus.musculus)
```

How to assign actual gene names to our counts. Using an annotation package for mouse 
```{r}
res_exp$symbol <- mapIds(Mus.musculus, #which annotation package model to use
                     keys=row.names(res_exp), #using rownames
                     column="SYMBOL", #add a column of symbols associated with row.names arguments for the model
                     keytype="ENSEMBL", #our rownames are gene id's 
                     multiVals="first")
```

```{r}
res_exp$ONTOLOGY <- mapIds(Mus.musculus,
                     keys=row.names(res_exp),
                     column="ONTOLOGYALL",
                     keytype="ENSEMBL",
                     multiVals="first")
```

```{r}
res_exp$Genename <- mapIds(Mus.musculus,
                     keys=row.names(res_exp),
                     column="GENENAME", #now add a column for gene names or gene description
                     keytype="ENSEMBL",
                     multiVals="first")
```


```{r}
res_exp$ensembl <- mapIds(Mus.musculus,
                     keys=row.names(res_exp),
                     column="ENSEMBL", #now add a column for gene names or gene description
                     keytype="ENSEMBL",
                     multiVals="first")
```
```{r}
res_exp$ENTREZID <- mapIds(Mus.musculus,
                     keys=row.names(res_exp),
                     column="ENTREZID", #now add a column for gene names or gene description
                     keytype="ENSEMBL",
                     multiVals="first")
```


#***Plotting results***
Heatmaps are a good way to visualize the most significant genes
```{r}
mat <- assay(vsd)[ head(order(res_exp$padj),
                            30), #top 30 genes 
                       ]
#mat <- mat - rowMeans(mat)
df <- as.data.frame(colData(vsd)[,
                                     "exposure"])
rownames(df)<-colnames(mat)
colnames(df)[1]<- "Exposure"
```

```{r}
tiff(file = "Figure/Male_Dorsal_Hippocampus_heatmap.tiff", units="in", width=8, height=5, res=1000)
heatmap<-pheatmap(mat, 
         annotation_col=df,
         labels_row = res_exp$symbol,
         fontsize = 6,
         scale = "row",
         show_colnames = F,
         cluster_cols = TRUE,
         main = "Male Dorsal Hippocampus Heatmap")
heatmap
dev.off()
heatmap
```

#***volcano plots***
```{r}
tiff(file = 'Figure/volcano/Male_Dorsal Hippocampus_volcano_plot.tiff', units="in", width=8, height=5, res=1000)
M_vol1<- EnhancedVolcano(res_exp,
    lab = res_exp$symbol,
    x = 'log2FoldChange',
    y = 'pvalue',
    title = "Male Dorsal Hippocampus Volcano Plot",
    subtitle = "",
    ylim = c(0, -log10(10e-8)),
    xlim = c(log2(10e-3), -log2(10e-3)),
    FCcutoff = 1.0,
    pCutoff = 0.05,
    labSize = 3.0,
    colAlpha = 1,
    legendLabels=c('Not sig.','Log (base 2) FC','p-value',
      'p-value & Log (base 2) FC'),
    legendPosition = 'right',
    legendLabSize = 12,
    legendIconSize = 3.0,
    gridlines.major = FALSE,
    gridlines.minor = FALSE,
    border = 'full',
    borderWidth = 0.5,
    borderColour = 'black')
M_vol1
dev.off()

M_vol1
```
#Export TSV results

```{r}
exp_res <- as.data.frame(res_exp)
write_tsv(exp_res, file="male_hippocampus_results.tsv")
write_csv(exp_res, file="male_hippocampus_results.csv")
write_csv(as.data.frame(assay(M_se)), file= "male_hippocampus_counts.csv")
```


#REPEAT FOR FEMALES
```{r}
se1<- se[,se$tissue_type == "Dorsal Hippocampus"] ##change tissue type here

#se1<-se1[ , !se1$bam_ID == c("Li21.bam")] #remove specific outliers here

se1 <- se1[rowSums(assay(se1)) >= 10, ]

F_se<-se1[,se1$sex == "female"]

#test<-M_se #[ , !se1$animal_ID ==] #remove specific outliers here
#colData(test)

se1$group<-paste(se1$exposure, se1$sex, sep = "_") #create a group variable for easy comparison
se1$group

#se1 <- se1[ , order( colnames( se1 ) ) ]
#head(assay(se1))
#colData(se1)

se1$group<-factor(se1$group)
se1$group
dds <- DESeqDataSet(F_se, 
                    design = ~ exposure) #comparison done between high treatment and vehicle?
```


#***Exploratory analysis and visualization***
*transformations*
The variance stabilizing transformation (VST) a goal of stablizing the variance across the range of values. produce log2-like values for high counts. 
```{r}
colData(dds)

vsd <- vst(dds)
```


Now for visuals, plot a principal components analysis (PCA) using  ggplot2
```{r}
data <- plotPCA(vsd, 
                  intgroup = 'exposure' , 
                  returnData=TRUE)
data
percentVar <- round(100 * attr(data, 
                                 "percentVar"))
```
#PCA Plot
```{r}
tiff(file = 'Figure/PCA/Female_Dorsal_Hippocampus_PCA_outlier_removed.tiff', units="in", width=8, height=5, res=1000)
PCA <-ggplot(data, aes(PC1, 
                   PC2, 
                   color=group,
                   shape= group)) + 
  stat_ellipse(aes(color= group))+
  ggtitle("Female Dorsal Hippocampus PCA")+
  geom_point(size=3) +
  scale_color_manual(values = c("#666666", "#FF66CC","#66CCCC", "#9966FF")) +
  scale_shape_manual(values = c(19, 15,17, 18)) +
  xlab(paste0("PC1: ",
              percentVar[1],
              "% variance")) +
  ylab(paste0("PC2: ",
              percentVar[2],
              "% variance"))+ 
  theme_bw() + 
  theme(panel.border =element_rect(colour = "black", fill=NA, size=1), 
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(), 
        axis.line = element_line(colour = "black"))
PCA
dev.off()
PCA
```

#***Differential Expression***
Lets run a differential expression pipeline using the raw counts

This function will print out a message for the various steps it performs. These are described in more detail in the manual page for DESeq, which can be accessed by typing ?DESeq. Briefly these are: the estimation of size factors (controlling for differences in the sequencing depth of the samples), the estimation of dispersion values for each gene, and fitting a generalized linear model.
```{r}
dds <- DESeq(dds)
resultsNames(dds)
```
Calling results without any arguments will extract the estimated log2 fold changes and p values for the last variable in the design formula. If there are more than 2 levels for this variable, results will extract the results table for a comparison of the last level over the first level. Treatment 6 vs vehicle
```{r}
res_exp<-results(dds, name = "exposure_PCB_vs_control")

res_exp

summary(res_exp)
```


#***Annotating results***
```{r}
columns(Mus.musculus)
```

How to assign actual gene names to our counts. Using an annotation package for mouse 
```{r}
res_exp$symbol <- mapIds(Mus.musculus, #which annotation package model to use
                     keys=row.names(res_exp), #using rownames
                     column="SYMBOL", #add a column of symbols associated with row.names arguments for the model
                     keytype="ENSEMBL", #our rownames are gene id's 
                     multiVals="first")
```

```{r}
res_exp$ONTOLOGY <- mapIds(Mus.musculus,
                     keys=row.names(res_exp),
                     column="ONTOLOGYALL",
                     keytype="ENSEMBL",
                     multiVals="first")
```

```{r}
res_exp$Genename <- mapIds(Mus.musculus,
                     keys=row.names(res_exp),
                     column="GENENAME", #now add a column for gene names or gene description
                     keytype="ENSEMBL",
                     multiVals="first")
```


```{r}
res_exp$ensembl <- mapIds(Mus.musculus,
                     keys=row.names(res_exp),
                     column="ENSEMBL", #now add a column for gene names or gene description
                     keytype="ENSEMBL",
                     multiVals="first")
```
```{r}
res_exp$ENTREZID <- mapIds(Mus.musculus,
                     keys=row.names(res_exp),
                     column="ENTREZID", #now add a column for gene names or gene description
                     keytype="ENSEMBL",
                     multiVals="first")
```


#***Plotting results***
Heatmaps are a good way to visualize the most significant genes
```{r}
mat <- assay(vsd)[ head(order(res_exp$padj),
                            30), #top 30 genes 
                       ]
#mat <- mat - rowMeans(mat)
df <- as.data.frame(colData(vsd)[,
                                     "exposure"])
rownames(df)<-colnames(mat)
colnames(df)[1]<- "Exposure"
```

```{r}
tiff(file = "Figure/Female_Dorsal_Hippocampus_heatmap.tiff", units="in", width=8, height=5, res=1000)
heatmap<-pheatmap(mat, 
         annotation_col=df,
         labels_row = res_exp$symbol,
         fontsize = 6,
         scale = "row",
         show_colnames = F,
         cluster_cols = TRUE,
         main = "Female Dorsal Hippocampus Heatmap")
heatmap
dev.off()
heatmap
```

#***volcano plots***
```{r}
tiff(file = 'Figure/volcano/Female_Dorsal_Hippocampus_volcano_plot.tiff', units="in", width=8, height=5, res=1000)
M_vol1<- EnhancedVolcano(res_exp,
    lab = res_exp$symbol,
    x = 'log2FoldChange',
    y = 'pvalue',
    title = "Female Dorsal Hippocampus Volcano Plot",
    subtitle = "",
    ylim = c(0, -log10(10e-8)),
    xlim = c(log2(10e-3), -log2(10e-3)),
    FCcutoff = 1.0,
    pCutoff = 0.05,
    labSize = 3.0,
    colAlpha = 1,
    legendLabels=c('Not sig.','Log (base 2) FC','p-value',
      'p-value & Log (base 2) FC'),
    legendPosition = 'right',
    legendLabSize = 12,
    legendIconSize = 3.0,
    gridlines.major = FALSE,
    gridlines.minor = FALSE,
    border = 'full',
    borderWidth = 0.5,
    borderColour = 'black')
M_vol1
dev.off()

M_vol1
```
#Export TSV results

```{r}
exp_res <- as.data.frame(res_exp)
write_tsv(exp_res, file="female_hippocampus_results.tsv")
write_csv(exp_res, file="female_hippocampus_results.csv")
write_csv(as.data.frame(assay(M_se)), file= "female_hippocampus_counts.csv")
```



